---
title: "stockstill_casey_quiz3_binf702"
output: html_notebook
author: "Casey Stockstill"
date: "2025-02-24"
output:
  word_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

Question 1 Restated: Among 64 TMB patients, the following 2×2 table compares the standard angiogram (gold standard) with a noninvasive test:

Noninvasive Test Negative Noninvasive Test Positive Angiogram Negative 21 8 Angiogram Positive 3 32 Using this data:

(a) Compute the sensitivity (i.e. the proportion of angiogram‐positive patients who test positive by the noninvasive method).
(b) Compute the specificity (i.e. the proportion of angiogram‐negative patients who test negative).
(c) Assuming a stroke prevalence of 20% among TMB patients, compute the positive predictive value (PV+).

```{r}
# Create the contingency table
ct <- matrix(c(21, 8, 3, 32), nrow = 2, byrow = TRUE)
rownames(ct) <- c("Angio_Negative", "Angio_Positive")
colnames(ct) <- c("Noninv_Negative", "Noninv_Positive")
print(ct)

# Compute sensitivity = TP/(TP+FN)
sensitivity <- ct["Angio_Positive", "Noninv_Positive"] / sum(ct["Angio_Positive",])
print(paste("Sensitivity =", round(sensitivity, 4)))

# Compute specificity = TN/(TN+FP)
specificity <- ct["Angio_Negative", "Noninv_Negative"] / sum(ct["Angio_Negative",])
print(paste("Specificity =", round(specificity, 4)))

# Given prevalence of stroke (among TMB patients) is 20%
prev <- 0.20
# Positive Predictive Value (PV+)
ppv <- (sensitivity * prev) / (sensitivity * prev + (1 - specificity) * (1 - prev))
print(paste("Positive Predictive Value (PV+) =", round(ppv, 4)))

```

```{r}
# Create a mosaic plot for the contingency table
mosaicplot(ct, main = "Angiogram vs Noninvasive Test", 
           color = c("lightblue", "pink"), xlab = "Noninvasive Test", 
           ylab = "Angiogram")

```

Question 2(a) Restated: For the Stanford Binet IQ test (mean = 100, SD = 16), Mensa membership requires scoring in the upper 2% of the population. Find the cutoff score (i.e. the 98th percentile).

R Code:

```{r}
# Calculate the 98th percentile (upper 2%) cutoff score
cutoff <- qnorm(0.98, mean = 100, sd = 16)
print(paste("Mensa cutoff score =", round(cutoff, 2)))

```

```{r}
# Plot the normal distribution for IQ scores
x <- seq(60, 160, length.out = 1000)
y <- dnorm(x, mean = 100, sd = 16)
plot(x, y, type = "l", main = "Stanford Binet IQ Distribution", 
     xlab = "IQ Score", ylab = "Density")
abline(v = cutoff, col = "red", lwd = 2, lty = 2)
text(cutoff + 2, max(y) * 0.5, labels = paste("Cutoff =", round(cutoff, 1)), 
     col = "red")

```

Question 2(b) Restated: Assume that the number of violent asthma attacks seen in an emergency ward follows a Poisson distribution with a 1‑day rate of λ = 2. What is the probability of observing 5 or more cases over a 2‑day period?

Note: Since the 1‑day parameter wasn’t explicitly given, we assume λ = 2 per day so that over 2 days the mean becomes 4.

R Code:

```{r}
# Define the one-day Poisson rate and compute the two-day rate
lambda_day <- 2
lambda_2day <- 2 * lambda_day  # for 2 days

# Calculate probability of observing 5 or more cases in 2 days
prob_5_or_more <- 1 - ppois(4, lambda = lambda_2day)
print(paste("Probability of 5 or more cases =", round(prob_5_or_more, 4)))

```

```{r}
# Compute Poisson probabilities for a range of counts
x_poisson <- 0:15
y_poisson <- dpois(x_poisson, lambda = lambda_2day)
barplot(y_poisson, names.arg = x_poisson, 
        main = "Poisson Distribution (2-day period, λ = 4)", 
        xlab = "Number of Cases", ylab = "Probability", 
        col = "lightgreen")
# Draw a vertical line at 5 cases to highlight the cutoff
abline(v = 5, col = "red", lwd = 2, lty = 2)

```

Question 2(c) Restated: The rate of myocardial infarction (MI) in 50–59‑year‑old disease‑free women is about 10 per 1,000 over 5 years (i.e. p = 0.01). In a sample of 1,000 postmenopausal women taking hormones, only 3 MIs were observed over 5 years. Using the binomial distribution (n = 1000, p = 0.01), assess whether 3 events represent an unusually small number (i.e. compute the probability of observing 3 or fewer events).

R Code:

```{r}
# Set parameters for the binomial distribution
n <- 1000
p_event <- 10 / 1000  # 0.01

# Calculate the probability of observing 3 or fewer MIs
p_small <- pbinom(3, size = n, prob = p_event)
print(paste("Probability of 3 or fewer events =", round(p_small, 4)))

```

```{r}
# Define a range for the number of events
x_binom <- 0:20
y_binom <- dbinom(x_binom, size = n, prob = p_event)
barplot(y_binom, names.arg = x_binom, 
        main = "Binomial Distribution of MI Events (n = 1000, p = 0.01)", 
        xlab = "Number of MI Events", ylab = "Probability", 
        col = "lightcoral")
# Highlight the bars corresponding to 0 to 3 events
rect(xleft = 0.5, ybottom = 0, xright = 4.5, ytop = max(y_binom)*1.1, 
     border = "blue", lwd = 2, lty = 2)

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
